---
output: pdf_document
---

```{r setup, include=FALSE}
# Initial setup
knitr::opts_chunk$set(echo = TRUE)

# Load libraries
library(xml2) # to parse XML
library(tidyverse) 
# a group of libraries that I will use manipulate strings and for the parse_english_nodes function

# Read in DDI codebook and  remove namespace
ddi <- read_xml("../data/DDI/4194_Cohorte_blessures_scolaires_en_2_5.xml")
ddi <- xml_ns_strip(ddi)

# function to parse multiple elements by attribute and return a single string
parse_nodes_for_attr <- function(nodes, attr, value, sep = ", ") {
  att <- as.vector(xml_attr(nodes, attr))
  text <- as.vector(xml_text(nodes))
  return <- data.frame(att = att, text = text)
  return <- return %>% filter(att == value & text != "") # the pipe (%>%) and filter are from tidyverse
  
  # using the str_replace function to remove extraneous punctuation at the end of the values. I'm using a regular expression for the pattern and I'm looking for zero or more(*) of the symbols in the brackets (":", ".", and " "). The $ signifies the end of the string. It's being replaced by nothing so that deletes the characters 
  return$text <- str_replace(return$text, "[:. ]*$", "") 
  
  # concatenating all the values and returning a single string
  str_c(return$text, collapse = sep) 
}
```

```{r, include = FALSE}
# As a guide, anything within three backticks following by {r} is r code. Also any statement within backticks like `r ...` is inline r code. Anything starting with a backslash is LaTeX. I tried to avoid using LaTeX as much as possible. 

# I'm only using the custom function I wrote above and function from xml2 to parse the XML for elements by attribute. When an element is not accessible by using the custom function, I will look for single elements using xml_find_all(ddi, "//element[#]). This is because there's not an attribute to use to pinpoint the element for the text that's to be included (or the same attribute is used multiple times). 

# The markdown I'm using consists of headers (#, ##, ###), formatting (** for bold & * for italic), block quotes which use the greater than sign (>), tables which use pipes and dashes (|---|---|), and links which use brackets and parenthesis to hyperlink text [text](hyperlink)

# I find it useful to have the XML file open which can be done in the viewer in RStudio if you install the htmltidy library and use htmltidy::xml_view(ddi)
```

\vspace{1in}

# `r xml_text(xml_find_first(ddi, "stdyDscr//citation//titlStmt//titl"))` / *`r xml_text(xml_find_first(ddi, "stdyDscr//citation//titlStmt//parTitl"))`*

Data Collectors: `r xml_text(xml_find_all(ddi, "//dataCollector[1]"))` (`r xml_attr(xml_find_all(ddi, "//dataCollector[1]"), "affiliation")`) & `r xml_text(xml_find_all(ddi, "//dataCollector[2]"))` (`r xml_attr(xml_find_all(ddi, "//dataCollector[2]"), "affiliation")`) 

\vspace{0.5in}

## Abstract

|  |  |
|---------|--------|
|`r parse_nodes_for_attr(xml_find_all(ddi, "//abstract"), "lang", "FR")`| `r parse_nodes_for_attr(xml_find_all(ddi, "//abstract"), "lang", "EN")`|

\vfill

**Keywords** 

|  |  |
|---------|--------|
|`r parse_nodes_for_attr(xml_find_all(ddi, "//keyword"), "lang", "FR")`|`r parse_nodes_for_attr(xml_find_all(ddi, "//keyword"), "lang", "EN")`

**Related publications**: [`r xml_text(xml_find_first(ddi, "//relPubl"))`](`r xml_text(xml_find_first(ddi, "//relPubl"))`)

\newpage

# Universe

|  |  |
|--------|---------|
|`r parse_nodes_for_attr(xml_find_all(ddi, "//universe"), "clusion", "I", sep = " | ")`|

`r xml_find_all(ddi, "//universe[5]")` / `r xml_find_all(ddi, "//universe[6]")`

**Type of study**: [`r xml_text(xml_find_first(ddi, "//timeMeth"))`](`r xml_attr(xml_find_first(ddi, "//timeMeth"), "method")`)

---

## Study population

`r xml_find_all(ddi, "//universe[3]")` / `r xml_find_all(ddi, "//universe[4]")`

> *Type de population/Population type*: `r parse_nodes_for_attr(xml_find_all(ddi, "//universe"), "level", "Type de population")` / `r parse_nodes_for_attr(xml_find_all(ddi, "//universe"), "level", "Population type")`

> *Tranche d'âge/Age range*: `r parse_nodes_for_attr(xml_find_all(ddi, "//universe"), "level", "Tranche d'âge")` / `r parse_nodes_for_attr(xml_find_all(ddi, "//universe"), "level", "Age range")`

> *Sexe/Sex*: `r parse_nodes_for_attr(xml_find_all(ddi, "//universe"), "level", "Sexe")` / `r parse_nodes_for_attr(xml_find_all(ddi, "//universe"), "level", "Sex")`

\newpage

# Data Collection

**Collection Dates**: `r parse_nodes_for_attr(xml_find_all(ddi, "//collDate"), "event", "start")`-`r parse_nodes_for_attr(xml_find_all(ddi, "//collDate"), "event", "end")`

**Collection Location:** `r xml_text(xml_find_all(ddi, "//nation"))`, `r xml_text(xml_find_first(ddi, "//geogCover"))`

---

## Sampling procedure

|   |   |
|---|---|
|`r parse_nodes_for_attr(xml_find_all(ddi, "//sampProc"), "lang", "FR")`| `r parse_nodes_for_attr(xml_find_all(ddi, "//sampProc"), "lang", "EN")`|

## Collection frequency

|   |   |
|---|---|
|`r parse_nodes_for_attr(xml_find_all(ddi, "//frequenc"), "lang", "FR")`| `r parse_nodes_for_attr(xml_find_all(ddi, "//frequenc"), "lang", "EN")`|

## Mode of collection

|   |   |
|---|---|
|`r xml_find_all(ddi, "//collMode[1]")`| `r xml_find_all(ddi, "//collMode[2]")`|

## Kinds of data collected

|   |   |
|---|---|
|`r parse_nodes_for_attr(xml_find_all(ddi, "//dataKind"), "lang", "EN", "; ")`|`r parse_nodes_for_attr(xml_find_all(ddi, "//dataKind"), "lang", "FR", "; ")`|
